Ornaments as indicators of social changes in northeastern Taiwan before and after European colonial period

### 2019-12-15 analysis/paper/paper.Rmd
remove period from fig captions, run all the code chuck

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Liu2011; @Chen2017]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeast Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other indigenous groups, Chinese merchants, and Europeans, via sea.

@@

Ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, are considered to have been imported from other regions. This is because of to a lack of archaeological evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition of glass beads from this region shows a high content of lead and, toghether with the winding/folding technique, these details suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common components of metal ornaments are brass and copper, with a small number made from lead and tin that indicates multiple origins that include Southeast Asia [@Chen2011]. There is no direct evidence showing European delivery of beads, however, a large amount of the glass beads containing gold foil (hereafter, gold-foil beads) at Kiwulan might have been introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, as part of the trading route of the Spanish between 16-19th century [@Wang2007]. Both archaeological evidence and historical records indicate northeastern Taiwan was involved inregional networks with East Asia in the late Iron age. These included Chinese merchants trading metal items, clothes, and beads with local indigenous people in Taiwan in exchange for local resources. The foreign-made large dark brown glazed stoneware jars frequently found in European shipwrecks were also commonly found from many sites in Taiwan, suggesting direct or indirect interactions. Despite the Chinese origin of some ornaments at Kiwulan, there is compelling evidence that a large amount of ornaments found at 17th century sites resulted from European colonial and economic activities in the region.

@@

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

```{r C14-data, fig.cap= "Summed probability distributions for dates from Kiwulan. The dark line represents the summed probabilities of all radiocarbon ages, and the grey lines in the background are the probabilities of individual ages. Grey rectangles indicate the approximate chronology of the major archaeological components of the deposit. For the upper component, the blue line indicates the start of European presence, while the red line is the Chinese presence. Ages calibrated with the Bchron package (Parnell et al. 2008).", cache=TRUE}

@@

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-c14-ages-distribution.jpg"))

```

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: bells, C: glass beads and gold-foil beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007).", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods."}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period_clean %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 10)

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, carnelian beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

```{r plot-all-densities, fig.cap= "Kernel density map for ornaments by periods. Used the bandwidth based on Silverman (1986)'s rule of thumb."}
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)
jpeg(
  here("analysis", "figures",
       "plot-kde-maps.jpg"), 
    width = 135,
    height = 40,
    units = "mm",
  res = 300)
par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)
kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
invisible(dev.off())
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.jpg"))
```

@@

```{r hypo-test-distribution, fig.cap= "Histograms of simulated ANN values from 1000 simulations for three time periods. X-axis values represent ANN expected value under a completely random process resulted from simulated pattern. Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value."}

>>>>

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Liu2011; @Chen2017]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeastern Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other indigenous groups, Chinese merchants, and Europeans, via sea.

@@

Ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, are considered to have been imported from other regions. This is because of to a lack of archaeological evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition of glass beads from this region shows a high content of lead and, toghether with the winding/folding technique, these details suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common components of metal ornaments are brass and copper, with a small number made from lead and tin that indicates multiple origins that include Southeast Asia [@Chen2011]. There is no direct evidence showing European delivery of beads, however, a large amount of the glass beads containing gold foil (hereafter, gold-foil beads) at Kiwulan might have been introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, as part of the trading route of the Spanish between 16-19th century [@Wang2007]. Both archaeological evidence and historical records indicate northeastern Taiwan was involved in regional networks with East Asia in the late Iron age. These included Chinese merchants trading metal items, clothes, and beads with local indigenous people in Taiwan in exchange for local resources. The foreign-made large dark brown glazed stoneware jars frequently found in European shipwrecks were also commonly found from many sites in Taiwan, suggesting direct or indirect interactions. Despite the Chinese origin of some ornaments at Kiwulan, there is compelling evidence that a large amount of ornaments found at 17th century sites resulted from European colonial and economic activities in the region.

@@

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

```{r C14-data, fig.cap= "Summed probability distributions for dates from Kiwulan. The dark line represents the summed probabilities of all radiocarbon ages, and the grey lines in the background are the probabilities of individual ages. Grey rectangles indicate the approximate chronology of the major archaeological components of the deposit. For the upper component, the blue line indicates the start of European presence, while the red line is the Chinese presence. Ages calibrated with the Bchron package (Parnell et al. 2008)", cache = TRUE}

@@

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-c14-ages-distribution.jpg"))
```

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: bells, C: glass beads and gold-foil beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007)", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods"}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period_clean %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 10)

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, carnelian beads, bells, and glass beads"}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

```{r plot-all-densities, fig.cap= "Kernel density map for ornaments by periods. Used the bandwidth based on Silverman (1986)'s rule of thumb"}
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)
jpeg(
  here("analysis", "figures",
       "plot-kde-maps.jpg"), 
    width = 135,
    height = 40,
    units = "mm",
  res = 300)
par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)
kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
invisible(dev.off())
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.jpg"))
```

@@

```{r hypo-test-distribution, fig.cap= "Histograms of simulated ANN values from 1000 simulations for three time periods. X-axis values represent ANN expected value under a completely random process resulted from simulated pattern. Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}
